---
title: "Project_RBI"
author: "Aashish Sukhija & Eliot Lepetit"
format: 
  html: default
  pdf: default
editor: 
  markdown: 
    wrap: sentence
---

## Import datasets

```{r}
#| message: false
#| warning: false
here::i_am("Project_RBI.Rproj")
library(here)       
library(readxl)     
library(tidyr)  
library(stringr)
library(dplyr)
library(vroom)
library(readr)
library(knitr)
library(ggplot2)
library(janitor)
library(purrr)
theme_set(theme_bw())
```

```{r}
suppressMessages({
Literacy_data <- read.csv(here("data", "State-wise_Literacy_Rate.csv"))
Birth_data <- read.csv(here("data", "State-wise_Birth_Rate.csv"))
Sex_data <- read.csv(here("data", "State-wise_Sex_Ratio.csv"))
Death_data <- read.csv(here("data", "State-wise_Death_Rate.csv"))
Pop_data <- read.csv(here("data", "State-wise_Total_Population.csv"))
IMR_data <- read.csv(here("data", "State-wise_Infant_Mortality_Rate.csv"))
TFR_data <- read.csv(here("data", "State-wise_Total_Fertility_Rate.csv"))
Life_data <- read.csv(here("data", "State-wise_Life_Expectancy.csv"))
Health_Exp_data <- read.csv(here("data", "State-wise_Public_Expenditure_on_Health.csv"))
NSDP1_data <- read.csv(here("data", "Net_State_Domestic_Product_1.csv"))
NSDP2_data <- read.csv(here("data", "Net_State_Domestic_Product_2.csv"))
NSDP3_data <- read.csv(here("data", "Net_State_Domestic_Product_3.csv"))
})
```

## Description of the datasets

We now describes the data files used in this analysis, their sources, and the processing steps.

-   GitHub Project Link to clone is : [Github_link](https://github.com/ELiot-Lepetit/Project_RBI.git)

-   Data Sources:

1.  [Handbook of Statistics on Indian States](https://www.rbi.org.in/Scripts/AnnualPublications.aspx?head=Handbook+of+Statistics+on+Indian+States)

2.  [Net State Domestic Product - State-wise (At Constant Prices)](https://www.rbi.org.in/Scripts/PublicationsView.aspx?i%20d=22480)

### Instruction to reproduce the data selection

1.  Click on the above links to reach the official website of India’s Central Bank, which provides the “Database on Indian Economy” and Net State Domestic Product state-wise for secondary research.
2.  Select the Social-demographic indicators for the year 2022-23, including Birth rate, Death rate, Population, Literacy rate, Infant Mortality rate, etc.
3.  Download the respective files as Excel spreadsheets from both the links.
4.  Convert the Excel files into CSV format for better data analysis of large datasets and to easily track changes in GitHub.

Note: We have already provided a zip file named `data.zip` in our GitHub project, which contains all the CSV files.

-   A short description:

The data for this project is derived from resources published by the Reserve Bank of India (RBI), specifically curated to provide insights into the economic and socio-demographic profiles of 28 states and 8 Union territories in the country.
These datasets serve as reliable and authoritative references for our research intent to analyze regional economic growth and its correlation with socio-demographic indicators.

1.  Socio-Demographic Indicators

The socio-demographic data is obtained from the Handbook of Statistics on Indian States for year 2022-2023, an annual publication by the Reserve Bank of India.
This dataset offers a comprehensive overview of the demographic, health, and social characteristics of Indian states, which are essential for understanding the human development aspects of economic growth.

The key indicators included are:

-   **State-wise Literacy Rate**: Proportion of the population aged seven and above who can read and write with understanding.
-   **State-wise Birth Rate**: Number of live births per thousand people annually.
-   **State-wise Death Rate**: Number of deaths per thousand people annually.
-   **State-wise Sex Ratio**: Number of females per thousand males.
-   **State-wise Total Population**: Total population of each state based on census and projections.
-   **State-wise Infant Mortality Rate (IMR)**: Number of infant deaths (under one year of age) per thousand live births.
-   **State-wise Total Fertility Rate (TFR)**: Average number of children a woman is expected to have during her lifetime.
-   **State-wise Life Expectancy**: Average number of years a person is expected to live at birth.
-   **State-wise Public Expenditure on Health**: State-level spending on public healthcare infrastructure and services.

These indicators provide a socio-demographic lens for evaluating the quality of life and human development across states.
The data enables an exploration of how socio-economic factors, such as education, health, and fertility, interact with economic growth, helping to explain variations in Net State Domestic Product.
We hope to take it a step forward by comparing states who are at the forefront and compare their Net State Domestic product.
We also hope to compare states by the geographical region they occupy in the country and try to find if they share the similar growth in NSDP over the years.

2.  Net State Domestic Product (NSDP) - State-wise (At Constant Prices)

The Net State Domestic Product (NSDP) data is sourced from the RBI’s State-wise Economic Indicators.
This dataset provides state-wise estimates of NSDP at constant prices, reflecting the total economic output of each state after accounting for depreciation (consumption of fixed capital).
NSDP is a crucial metric for assessing the economic performance of individual states, as it indicates the value of goods and services produced within a state’s geographical boundaries during a given period.

We eliminate the impact of inflation through using constant prices, allowing for meaningful comparisons of economic growth over time and across states.
This dataset is organized by fiscal years and further disaggregated into sectoral contributions, including agriculture, industry, and services.
It will help us as potential researchers to identify the dominant economic activities in each state and evaluate trends in economic growth.

The NSDP data is particularly valuable for understanding regional disparities, exploring growth trajectories, and assessing the impact of policy interventions.
By focusing on state-level economic activity, this resource provides insights into the structural differences that drive variations in development outcomes across India.

**Integration and Use in Analysis**

All indicators in the datasets complement each other by linking economic performance (NSDP) with socio-demographic factors.
While the NSDP data offers insights into the scale and composition of economic activity, the socio-demographic indicators reveal the underlying human capital and health factors influencing economic growth.
Together, they enable a multi-dimensional analysis of regional disparities, allowing researchers to examine how socio-demographic conditions impact economic outcomes and vice versa.

These datasets are well-structured, publicly available, and authoritative, making them ideal for analyzing the interplay between economic performance and human development across Indian states along with the adversities.

## Basic Information about the datasets

The Datasets table represents the 11 raw data sets before the cleaning and merging which we will use for visualization and interpretation of our research questions/interests.
The number of rows depicts the the number of states and union territories and the number of columns represent the availability of data with respect to the indicators and Net State Domestic Product by year.

```{r}
datasets <- list(
  Literacy_data, Birth_data, Sex_data, Death_data, Pop_data, 
  IMR_data, Life_data, Health_Exp_data, NSDP1_data, NSDP2_data, NSDP3_data
)

# Dataset names
dataset_names <- c(
  "Literacy_data", "Birth_data", "Sex_data", "Death_data", "Pop_data", 
  "IMR_data", "Life_data", "Health_Exp_data", "NSDP1_data", "NSDP2_data", "NSDP3_data"
)

# Create a tibble with Rows and Columns calculated dynamically
data_info <- tibble(
  Datasets = dataset_names,
  `States/Union Territories` = sapply(datasets, nrow),
  Years = sapply(datasets, ncol)
)

# Generate the Markdown table using knitr::kable
kable(data_info, caption = "Summary of Datasets")
```

## Cleaning data

In this part we removed the unused extra rows which came from the conversion from excel files to csv files.
We also ensured to standardize the format of each database to facilitate merging.

### Cleaning Social and Demographic Indicators datasets to prepare for merging with Net State Domestic Product Dataset

#### State-wise Birth Rate

```{r}
#| message: false
#| warning: false
Birth_data <- Birth_data %>%  rename(State = State.Union.Territory ) %>%
      rename_with(~ sub("^X", "", .)) %>%
    mutate(across(-State, as.numeric)) %>% 
    slice(-c(39, 40)) %>%
  mutate(across(where(is.numeric), ~ round(., 1))) #rounding the data to one decimal place
```

#### State-wise Death Rate

```{r}
#| message: false
#| warning: false
Death_data <- Death_data %>%  rename(State = `State.Union.Territory`) %>%
      rename_with(~ sub("^X", "", .)) %>%
    mutate(across(-State, as.numeric)) %>%
  slice(-c(39, 40)) %>%
   mutate(across(where(is.numeric), ~ round(., 1))) #rounding the data to one decimal place
```

#### State-wise Public Expenditure on Health

```{r}
#| message: false
#| warning: false
Health_Exp_data <- Health_Exp_data %>%
  rename(State = `State.Union.Territory`) %>%                   
  rename_with(~ gsub("^X|\\..*", "", .)) %>%                     
  mutate(across(-State, as.numeric)) %>%                        
  mutate(across(where(is.numeric), ~ round(., 1))) %>%           
  slice(-c(37, 38))                                             
```

#### State-wise Literacy Rate

```{r}
#| message: false
#| warning: false
Literacy_data <- Literacy_data %>% rename(State = `State.Union.Territory`) %>%
        rename_with(~ sub("^X", "", .)) %>%
  mutate(across(-State, as.numeric)) %>%
  mutate(across(where(is.numeric), ~ round(., 1))) %>%
  slice(-c(37:40))
```

#### State-wise Total population

```{r}
#| message: false
#| warning: false
Pop_data <- Pop_data %>% rename(State = `State.Union.Territory`) %>%
        rename_with(~ sub("^X", "", .)) %>%
  slice(-c(37:43)) %>%
  mutate(across(everything(), ~ ifelse(. == "-", NA, .)))
```

#### State-wise Sex Ratio

```{r}
#| message: false
#| warning: false
Sex_data <- Sex_data %>% rename(State = `State.Union.Territory`) %>%
      rename_with(~ sub("^X", "", .)) %>%
  slice(-c(37:39))  %>%
  mutate(across(everything(), ~ ifelse(. == "-", NA, .))) %>%
  select(-`1951`)  
```

#### State-wise Total Fertility Rate

```{r}
#| message: false
#| warning: false
TFR_data <- TFR_data %>%  rename(State = `Year`) %>%
    rename_with(~ sub("^X", "", .)) %>%
    mutate(across(-State, as.numeric)) %>%
  slice(-c(24:26)) %>%
   mutate(across(where(is.numeric), ~ round(., 1)))
```

#### Infant mortality Rate

```{r}
#| message: false
#| warning: false
IMR_data <- IMR_data %>%
  row_to_names(row_number = 1) %>%
  slice(-c(38:41)) %>%
  rename(State = `State/Union Territory`) %>%
  mutate(across(-State, ~ as.numeric(gsub("\\*", "", .))))
```

### Merging and cleaning the NSDP datasets to have all states in one database

The excel file contained 3 sheets which we will be merging and then pivoting to have one observation for each state (stored in a varibale named "State").

```{r}
#| message: false
#| warning: false
NSDP_merged <- NSDP1_data %>%
  full_join(NSDP2_data, by = "Year") %>%
  full_join(NSDP3_data, by = "Year")

# Removing unused rows (we keep the data only for years after 2011)
NSDP_merged <- NSDP_merged[c(11:16, 25:29, 38:44, 53:65), ] 

# Transforming missing values character as "NAs" and changing the dates format
NSDP_merged <- NSDP_merged %>% 
  mutate(across(everything(), ~ ifelse(. == "-", NA, .))) %>%
  mutate(Year = as.numeric(sub("-.*", "", Year)))

# We pivot the database to have a variable "State" in order to merge with other datasets
NSDP_long <- NSDP_merged %>%
  mutate(across(-Year, as.character)) %>%
  pivot_longer(
    cols = -Year,         
    names_to = "State",    
    values_to = "Value"     
  ) 

# Clean and arrange the State variable in alphabetical order and replace specific names
NSDP_clean <- NSDP_long %>%
  pivot_wider(
    names_from = Year,     
    values_from = Value
  )  %>%
  mutate(across(-State, as.numeric)) %>%
  mutate(
    State = gsub("\\.+", " ", State),                        # Replace "." with " "
    State = case_when(
      State == "Andaman Nicobar Islands" ~ "Andaman & Nicobar Islands", # Fix specific name
      State == "Jammu Kashmir" ~ "Jammu & Kashmir",                     # Fix Jammu Kashmir
      TRUE ~ State                                                    # Keep other names as is
    )
  ) %>%
  arrange(State)  # Sort by State in alphabetical order

rm(NSDP1_data, NSDP2_data, NSDP3_data)

```

#### Graphical representation of the NSDP (per state) variable.

For visibility, we only display the first five states: `r paste(NSDP_clean$State[1:5], collapse = ", ")`.

```{r}
#| message: false
#| warning: false
NSDP_plot <- NSDP_clean %>%
  slice(1:5) %>%
  pivot_longer(
    cols = -State,          
    names_to = "Year",       
    values_to = "Value"
  ) %>%
    mutate(Year = as.numeric(Year))

library(scales)
ggplot(NSDP_plot, aes(x = Year, y = Value, color = State)) +
  geom_line(size = 1) +                
  labs(
    title = "Trends in NSDP by State/Union Territory",
    x = "Year",
    y = "NSDP Value (in Crores)"
  ) +
  scale_y_continuous(
    breaks = seq(0, max(NSDP_plot$Value, na.rm = TRUE), by = 100000),  
    labels = label_number(scale_cut = cut_short_scale())              
  ) +
  scale_x_continuous(
    breaks = seq(min(NSDP_plot$Year, na.rm = TRUE), max(NSDP_plot$Year, na.rm = TRUE), by = 2) 
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1), 
    panel.grid.major.y = element_line(color = "gray", linetype = "dotted"))  
```

## Description of Research Question

-   Research Question:- Investigating the Interplay Between Socio-Demographic Indicators and Net State Domestic Product (NSDP) Across Indian States

-   Why these Datasets:- The relationship between socio-demographic indicators and economic performance analyze the complexities of regional development in a diverse country like India.
    This research seeks to investigate how socio-demographic factors, such as literacy rate, health expenditure, birth rate, and other indicators, interact with the Net State Domestic Product (NSDP) to shape the economic outcomes of Indian states.
    By linking these indicators, we aim to uncover patterns, disparities, and trends that explain variations in development outcomes across different states of India.

The NSDP data reflects the capacity of states to generate economic value, adjusted for inflation, allowing for meaningful comparisons over time.
On the other hand, socio-demographic indicators reveal the underlying health and human capital conditions that can either drive or hinder economic growth.
Together, these datasets provide a robust foundation for analysis of regional disparities, enabling us to examine how social conditions influence economic outcomes and vice versa.

-   Objectives of the Research:-
    1.  Understanding Regional Disparities: • Analyze differences in NSDP values across states and identify the factors contributing to these disparities. • Explore why some states (e.g., Maharashtra, Tamil Nadu) consistently outperform others (e.g., Bihar, Odisha) in terms of economic growth.
    2.  Exploring the Role of Socio-Demographic Indicators: • Investigate the correlation between NSDP and socio-demographic variables like literacy rates, health expenditure, and birth/death rates. • Examine whether higher investments in human capital and health infrastructure translate to better economic performance.
    3.  Analyzing Growth Trajectories: • Evaluate how NSDP and socio-demographic indicators have evolved over time, identifying trends and growth trajectories for individual states.
    4.  Regional and Cultural Comparisons: • Group states by geographic regions (e.g., northern, southern, northeastern) to understand how shared cultural and geographical factors influence economic and social outcomes. • Assess whether regional similarities (e.g., among northeastern states) lead to comparable growth patterns in NSDP and socio-demographic indicators.
-   Why this Research Questions:-

India’s diversity in economic and social conditions makes it an ideal case for studying the interplay between economic and demographic factors.
While high-performing states may benefit from robust health and education systems, underperforming states may struggle due to insufficient investments in human capital.
By focusing on these dynamics, this research can provide insights into the structural differences that drive variations in state-level development.

And, also understanding the impact of cultural and geographic differences adds a unique layer to the analysis.
For instance, states in the northeast may face challenges due to geographic isolation, while southern states may excel due to better health and education policies.

-   Conclusion:- This project has the potential to serve as a valuable resource for understanding the factors shaping economic performance in developing nations, offering lessons for addressing disparities and fostering equitable development. By linking social and economic datasets, this study not only explains past growth patterns but also informs future strategies for sustainable development.

## Merging datasets

```{r}
datasets <- list(
  Literacy_data, Birth_data, Sex_data, Death_data, Pop_data, Health_Exp_data, NSDP_clean, TFR_data, IMR_data
)

# Names of the datasets
dataset_names <- c(
  "Literacy Rate", "Birth Rate", "Sex Ratio", "Death Rate", "Population", "Health Expenditure", 
  "NSDP", "Total Fertility Rate", "Infant Mortality Rate"
)

# Preprocessing function
preprocess_dataset <- function(data) {
  data %>%
    {if (any(grepl("State", colnames(.)))) rename(., State = starts_with("State")) else .} %>% # Rename State if it exists
    rename_with(~ gsub("^X", "", .), -State) %>%    # Remove "X" prefix from year columns
    {if ("1951" %in% colnames(.)) select(., -`1951`) else .} %>% # Remove "1951" if it exists
    mutate(across(-State, as.character))           # Ensure all year columns are character
}

# Apply preprocessing to all datasets
datasets <- map(datasets, preprocess_dataset)

# Convert each dataset to long format
long_datasets <- map2(datasets, dataset_names, function(data, name) {
  data %>%
    pivot_longer(
      cols = -State,                # Keep "State" column
      names_to = "Year",            # Create "Year" column from column names
      values_to = name              # Use dataset name as the value column
    ) %>%
    mutate(Year = as.numeric(Year)) # Convert Year column back to numeric
})

# Merge all long-format datasets by "State" and "Year"
merged_data <- reduce(long_datasets, full_join, by = c("State", "Year"))
```

### 2019 Analysis (Hyp 1)

```{r}
merged_data_2019 <- merged_data %>%
  filter(Year == 2019) %>%
  filter(!is.na(NSDP)) #We remove the NAs for NSDP
```

### 2011 Analysis (Hyp 2)

```{r}
merged_data_2011 <- merged_data %>%
  filter(Year == 2011) %>%
  filter(!is.na(NSDP)) #We remove the NAs for NSDP
```


## Merged table display
```{r}
# Show the first 5 rows and the last 5 rows
table_to_present <- merged_data %>%
  filter(State!= "ALL INDIA") %>%
  filter(Year %in% c(2011, 2019)) %>% # Filter for specific years
  {bind_rows(head(., 3), tail(., 3))} # Combine first 5 rows and last 5 rows

# Render the table
kable(table_to_present, caption = "Summary of the Merged Dataset", align = "c")
```





